PhD Thesis

### 2016-12-02 pldi14/sec-change-structures.tex
Prepare integration: fix conflicts with polytable

This definition is interesting because we \emph{can} define the
change structure, but we have issues with its elimination forms.
Take a pair of functions $p = (f, g)$ and its change
$dp = \mathit{inj}_3 db$ (with $db : \Delta b)$. How are we to
compute the derivative of its first projection, $fst' p dp$? This
must be the nil change of $f$, but this has a few issues:
\begin{itemize}
\item such a derivative is not self-maintainable because it
  requires knowing $f$, so it is inefficient in the framework
  present in this paper; however, we are addressing this issue.
\item The result is an inefficient nil change, because we do not
  use a computed derivative but simply
  $\Nil{f} = \Diff{f}{f} = \Lam{x}{\Lam{dx}{\Diff{\App{f}{\Update
          x {\D{x}}}}{\App{f}{x}}}}$. One can imagine a few ways
  to avoid that, but some require nontrivial changes to the
  framework. For instance, one could add derivatives as an
  additional observation of basic functions, so that whenever we
  have access to a function we also have access to its
  derivative. We leave investigation on this point for future
  work.
\item The semantics of a pair or of its two elements together
  should arguably be equivalent by general reasons --- simply
  because the environment is itself a sort of product (that's for
  instance guaranteed by categorical semantics). This suggests
  that, if we allow omitting changes for components of products,
  we should also allow omitting them for entries in the context.
\item It is currently sometimes hard to detect whether a change
  is nil or not. We should arguably add a suitable method to
  change structures (which is specified to give a conservative
  answer), and implement it for function changes through yet
  another observation.
\end{itemize}

>>>>

This definition is interesting because we \emph{can} define the
change structure, but we have issues with its elimination forms.
Take a pair of functions $p = (f, g)$ and its change
$dp = \mathit{inj}_3 db$ (with $db : \Delta b)$. How are we to
compute the derivative of its first projection, $fst' p dp$? This
must be the nil change of $f$, but this has a few issues:
\begin{itemize}
\item such a derivative is not self-maintainable because it
  requires knowing $f$, so it is inefficient in the framework
  present in this paper; however, we are addressing this issue.
\item The result is an inefficient nil change, because we do not
  use a computed derivative but simply
  $\NilC{f} = \Diff{f}{f} = \Lam{x}{\Lam{dx}{\Diff{\App{f}{\Update
          x {\D{x}}}}{\App{f}{x}}}}$. One can imagine a few ways
  to avoid that, but some require nontrivial changes to the
  framework. For instance, one could add derivatives as an
  additional observation of basic functions, so that whenever we
  have access to a function we also have access to its
  derivative. We leave investigation on this point for future
  work.
\item The semantics of a pair or of its two elements together
  should arguably be equivalent by general reasons --- simply
  because the environment is itself a sort of product (that's for
  instance guaranteed by categorical semantics). This suggests
  that, if we allow omitting changes for components of products,
  we should also allow omitting them for entries in the context.
\item It is currently sometimes hard to detect whether a change
  is nil or not. We should arguably add a suitable method to
  change structures (which is specified to give a conservative
  answer), and implement it for function changes through yet
  another observation.
\end{itemize}

### 2016-12-02 pldi14/sec-change-theory.tex
Integrate PLDI14 paper
Prepare integration: fix conflicts with polytable

\paragraph{Nil changes and derivatives}
A particularly important change is the \emph{nil change} of a value:
\begin{definition}[Nil change]
  \label{def:nil-change}
  Given a change structure $\ChangeStruct{V}$ and a value $v \in V$, the change
  $\Diff{v}{v}$ is the nil change for $v$.
  \[
    \Nil{v} = \Diff{v}{v} \qed
  \]
\end{definition}
The nil change for a value does indeed not change it.
\begin{lemma}[Behavior of $\NIL$]
  \label{thm:update-nil}
  Given a change structure $\ChangeStruct{V}$ and a value $v \in V$,
  $\Apply{\Nil{v}}{v} = v$.
\end{lemma}

@@

Applying a derivative to a value and its nil change gives a nil
change.%
\footnote{Post-print note: There's a small technical mistake in
  the following lemma. See \cref{sec:change-eq} for a corrected
  statement.}
%
\begin{lemma}[Behavior of derivatives on $\NIL$]
  \label{thm:deriv-nil}
  Given change structures $\ChangeStruct{A}$ and
  $\ChangeStruct{B}$, a function $f \in A \to B$, an element $a$
  of $A$, and the derivative $f'$ of $f$, we have
  $\App{\App{f'}{a}}{\Nil{a}} = \Nil{\App* f a}$.
\end{lemma}

@@

\input{sec-function-change}

>>>>

\paragraph{Nil changes and derivatives}
A particularly important change is the \emph{nil change} of a value:
\begin{definition}[Nil change]
  \label{def:nil-change}
  Given a change structure $\ChangeStruct{V}$ and a value $v \in V$, the change
  $\Diff{v}{v}$ is the nil change for $v$.
  \[
    \NilC{v} = \Diff{v}{v} \qed
  \]
\end{definition}
The nil change for a value does indeed not change it.
\begin{lemma}[Behavior of $\NILC$]
  \label{thm:update-nil}
  Given a change structure $\ChangeStruct{V}$ and a value $v \in V$,
  $\Apply{\NilC{v}}{v} = v$.
\end{lemma}

@@

Applying a derivative to a value and its nil change gives a nil
change.%
\footnote{Post-print note: There's a small technical mistake in
  the following lemma. See \cref{sec:change-eq} for a corrected
  statement.}
%
\begin{lemma}[Behavior of derivatives on $\NILC$]
  \label{thm:deriv-nil}
  Given change structures $\ChangeStruct{A}$ and
  $\ChangeStruct{B}$, a function $f \in A \to B$, an element $a$
  of $A$, and the derivative $f'$ of $f$, we have
  $\App{\App{f'}{a}}{\NilC{a}} = \NilC{\App* f a}$.
\end{lemma}

@@

\input{pldi14/sec-function-change}

### 2016-12-02 pldi14/sec-function-change.tex
Integrate PLDI14 paper
Prepare integration: fix conflicts with polytable

% moved here to avoid annoying out-of-order figures.
\input{fig-base-derive}

@@

\begin{definition}
  \label{def:function-changes:change}
  Given change structures $\ChangeStruct{A}$ and $\ChangeStruct{B}$ and
  $f \in A \to B$,
  the set $\Change[A \to B]{f}$ contains all binary functions $\D
  f$ such that
  \NewDocumentCommand{\TheNewValue}{}{\Upd*{a}}
  \begin{subdefinition}
    \item
      \label{def:function-changes:signature}
      $\App{\App{\D f}{a}}{\D a} \in \Change[B]{\App*{f}{a}}$ and
    \item
      \label{def:function-changes:validity}
      $\App{f}{a} \UPDATE \App{\App{\D f}{a}}{\D a} =
      {\App{f}{\TheNewValue}}
      \UPDATE
      \App{\App{\D f}{\TheNewValue}}{\Nil{\TheNewValue}}$
  \end{subdefinition}
  for all values $a \in A$ and corresponding changes $\D a \in
  \Change[A]{a}$.
\end{definition}

@@

In particular, different functions of the same type can have
different sets of changes. Consider two functions of type
$\Fun{\Bag S}{\Bag S}$.
\begin{align*}
\App{f}{x} & = \Empty & \App{\Var{id}}{x} & = x
\end{align*}
The set
$\Change[\Fun{\Bag S}{\Bag S}] f$ contains ``changes'' to $f$,
namely all binary bag functions $df$ satisfying
(b): $\D{f}~a~\D{a}=\D{f}\APP\Upd*{a}\APP\Nil{\Upd*{a}}= \D{f}~(\MERGE~a~\D{a})~\Empty$.
Such binary functions include
$\MERGE$ and all constant functions.

The set $\Change[\Fun{\Bag S}{\Bag S}]\Term{id}$ contains changes to $id$,
namely all binary bag functions $\D{id}$ satisfying
(b):
$\Term{id}\APP a \UPDATE \D{id} \APP a \APP \D{a} =
\Term{id} \APP \Upd*{a} \UPDATE \D{id} \APP \Upd*{a} \APP
\Nil{\Upd*{a}}$, which simplifies to
$\MERGE~a~(\D{id}~a~\D{a})=
\MERGE~(\MERGE~a~\D{a})~(\D{id}~(\MERGE~a~\D{a})~\Empty)$.
Neither $\MERGE$ nor any constant function is a change to
$\Term{id}$,
but the function
$
\D{id}~a~\D a = \Merge{\D a}{\Set{1,2}}
$ is.
\end{examples}

@@

% Maybe reduce subscripts here?
\begin{definition}[Operations on function changes]
  \label{def:function-changes:update}
  \label{def:function-changes:diff}
  Given change structures $\ChangeStruct{A}$ and $\ChangeStruct{B}$,
  the operations $\APPLY[A \to B]$ and $\DIFF[A \to B]$ are
  defined as follows.
  %
  \begin{alignat*}{5}
    &\App{(\Update[A \to B]{f&&}{\D f})}{&&v}
      && = \Update[B]{\App{f}{v}&&}{\App{\App{\D f}{v}}{\Nil[A]{v}}}\\
    &\App{\App{(\Diff[A \to B]{f_2&&}{f_1})}{&&v}}{\D v}
      && = \Diff[B]{\App{f_2}{\Update*[A]{v}{\D v}}&&}{\App{f_1}{v}}\qedAligned
  \end{alignat*}
\end{definition}

@@

  Let $a_1 \in A$ be an arbitrary value with a corresponding
  change $\D a \in \Change[A]{a}$, and let $a_2$ be
  $\Apply{\D a}{a_1}$, then
  \begin{align*}
  & \Apply[B]
      {\App{\App{\Diff*[A \to B]{f_2}{f_1}}{a_1}}{\D a}}
      {\App{f_1}{a_1}}\\
  & \quad = \Apply[B]
               {\Diff*[B]
                 {\App{f_2}{a_2}}
                 {\App{f_1}{a_1}}}
               {\App{f_1}{a_1}}\\
  & \quad = \App{f_2}{a_2}\\
  & \quad = \Apply[B]
              {\Diff*[B]
                {\App{f_2}{a_2}}
                {\App{f_1}{a_2}}}
              {\App{f_1}{a_2}}\\
  & \quad = \Apply[B]
              {\Diff*[B]
                {\App{f_2}{\Apply*{\Nil[B]{a_2}}{a_2}}}
                {\App{f_1}{a_2}}}
              {\App{f_1}{a_2}}\\
  & \quad = \Apply[B]
              {\App{\App{\Diff*[A \to B]{f_2}{f_1}}{a_2}}{\Nil{a_2}}}
              {\App{f_1}{a_2}}
  \end{align*}
  by
  \cref{def:function-changes:diff,def:update-diff,thm:update-nil}.
\end{optionalproof}

@@

  Let $f_1, f_2 \in A \to B$ be arbitrary functions. We show that
  $\Apply[A \to B]{\Diff*[A \to B]{f_2}{f_1}}{f_1}$ is
  extensionally equal to $f_2$. Let $a \in A$ be an arbitrary
  value, then
  \begin{align*}
    & \App{\Apply*[A \to B]{\Diff*[A \to B]{f_2}{f_1}}{f_1}}{a}\\
    & \quad = \Apply[B]
                {\App{\App{\Diff*[A \to B]{f_2}{f_1}}{a}}{\Nil[A]{a}}}
                {\App{f_1}{a}}\\
    & \quad = \Apply[B]
                {\Diff*[B]{\App{f_2}{\Apply*[A]{a}{\Nil[A]{a}}}}{\App{f_1}{a}}}
                {\App{f_1}{a}}\\
    & \quad = \Apply[B]
                {\Diff*[B]{\App{f_2}{a}}{\App{f_1}{a}}}
                {\App{f_1}{a}}\\
    & \quad = \App{f_2}{a}
  \end{align*}
  by the definitions of $\APPLY[A \to B]$ and $\DIFF[A \to B]$,
  \cref{thm:update-nil} for the change structure $\ChangeStruct{A}$ and
  \cref{def:update-diff} for the change structure $\ChangeStruct{B}$.
\end{optionalproof}

@@

  Let $f$, $a$, $\D f$ and $\D a$ be arbitrary, as in the statement. Then
  \begin{align*}
    & \App{\Apply*[A \to B]{\D f}{f}}{\Apply*[A]{\D a}{a}}\\
    & \quad = \Apply[B]{\App{\App{\D f}{\TheNewValue}}{\Nil{\TheNewValue}}}{\App{f}{\TheNewValue}}\\
    & \quad = \Apply[B]{\App{\App{\D f}{a}}{\D a}}{\App{f}{a}}
  \end{align*}
  by
  \cref{def:function-changes:update,def:function-changes:validity}
  as required.
\end{optionalproof}

@@

One can also compute a pointwise change from a function change:
%In particular, a pointwise change can be obtained from a function
%change by substituting to $da$ a nil change $\Nil{a}$. The result of
%$\App{\App{f'}{\Old{a}}}{\Nil{a}}$ is also a nil change (by
%\cref{thm:deriv-nil}), and $\New{a} = \Old{a}$, so we obtain:
\[
  \Update{\App {f} {a}}{\App{\App{\D f}{a}}{\Nil{a}}}
= \App{f}{a} \UPDATE \App{\nabla f}{a}
\]

%If we substitute $\App{\nabla f}{\New{a}}$ away in the equation before, we obtain the equality:
%\begin{align*}
%  & \Update{\App {\Old{f}} {\Old{a}}}{\App{\App{\D f}{\Old{a}}}{\D a}} \\
%= & \App{\Old{f}}{\New{a}} \UPDATE \App{\nabla f}{\New{a}} \\
%= & \App{\Old{f}}{\New{a}} \UPDATE \App{\App{\D f}{\New{a}}}{\Nil{\New{a}}}
%\end{align*}
%
%The above discussion was informal. To formalize it, we must
%proceed in the opposite way: we incorporate this equality in the
%definition of function changes, define $\UPDATE$ and $\DIFF$ for
%changes, and only then we can finally state and prove the
%incrementalization theorem, since the formal statement depends on
%the definition of change structures.

@@

\begin{theorem}[Nil changes are derivatives]
  \label{thm:nil-is-derivative}
  Given change structures $\ChangeStruct{A}$ and $\ChangeStruct{B}$ and a function $f \in A \to B$,
  the change $\Nil[A \to B]{f}$ is the derivative $f'$ of $f$.
\end{theorem}

\begin{optionalproof}
  Let $a \in A$ be an arbitrary value with a corresponding change
  $\D a \in \Change[A]{a}$. Then
  \begin{align*}
    & \App{f}{\Apply*[A]{\D a}{a}}\\
    & \quad = \App{\Apply*[A \to B]{\Nil[A \to B]{f}}{f}}{\Apply*[A]{\D a}{a}}\\
    & \quad = \Apply[B]{\App{\App{\Nil[A]{f}}{a}}{\D a}}{\App{f}{a}}
  \end{align*}
  holds by \cref{thm:update-nil,thm:incrementalization}, as
  required for derivatives by \cref{def:derivatives}.
\end{optionalproof}

@@

\input{fig-function-change}
\end{oldSec}

>>>>

% moved here to avoid annoying out-of-order figures.
\input{pldi14/fig-base-derive}

@@

\begin{definition}
  \label{def:function-changes:change}
  Given change structures $\ChangeStruct{A}$ and $\ChangeStruct{B}$ and
  $f \in A \to B$,
  the set $\Change[A \to B]{f}$ contains all binary functions $\D
  f$ such that
  \NewDocumentCommand{\TheNewValue}{}{\Upd*{a}}
  \begin{subdefinition}
    \item
      \label{def:function-changes:signature}
      $\App{\App{\D f}{a}}{\D a} \in \Change[B]{\App*{f}{a}}$ and
    \item
      \label{def:function-changes:validity}
      $\App{f}{a} \UPDATE \App{\App{\D f}{a}}{\D a} =
      {\App{f}{\TheNewValue}}
      \UPDATE
      \App{\App{\D f}{\TheNewValue}}{\NilC{\TheNewValue}}$
  \end{subdefinition}
  for all values $a \in A$ and corresponding changes $\D a \in
  \Change[A]{a}$.
\end{definition}

@@

In particular, different functions of the same type can have
different sets of changes. Consider two functions of type
$\Fun{\Bag S}{\Bag S}$.
\begin{align*}
\App{f}{x} & = \Empty & \App{\Var{id}}{x} & = x
\end{align*}
The set
$\Change[\Fun{\Bag S}{\Bag S}] f$ contains ``changes'' to $f$,
namely all binary bag functions $df$ satisfying
(b): $\D{f}~a~\D{a}=\D{f}\APP\Upd*{a}\APP\NilC{\Upd*{a}}= \D{f}~(\MERGE~a~\D{a})~\Empty$.
Such binary functions include
$\MERGE$ and all constant functions.

The set $\Change[\Fun{\Bag S}{\Bag S}]\Term{id}$ contains changes to $id$,
namely all binary bag functions $\D{id}$ satisfying
(b):
$\Term{id}\APP a \UPDATE \D{id} \APP a \APP \D{a} =
\Term{id} \APP \Upd*{a} \UPDATE \D{id} \APP \Upd*{a} \APP
\NilC{\Upd*{a}}$, which simplifies to
$\MERGE~a~(\D{id}~a~\D{a})=
\MERGE~(\MERGE~a~\D{a})~(\D{id}~(\MERGE~a~\D{a})~\Empty)$.
Neither $\MERGE$ nor any constant function is a change to
$\Term{id}$,
but the function
$
\D{id}~a~\D a = \Merge{\D a}{\Set{1,2}}
$ is.
\end{examples}

@@

% Maybe reduce subscripts here?
\begin{definition}[Operations on function changes]
  \label{def:function-changes:update}
  \label{def:function-changes:diff}
  Given change structures $\ChangeStruct{A}$ and $\ChangeStruct{B}$,
  the operations $\APPLY[A \to B]$ and $\DIFF[A \to B]$ are
  defined as follows.
  %
  \begin{alignat*}{5}
    &\App{(\Update[A \to B]{f&&}{\D f})}{&&v}
      && = \Update[B]{\App{f}{v}&&}{\App{\App{\D f}{v}}{\NilC[A]{v}}}\\
    &\App{\App{(\Diff[A \to B]{f_2&&}{f_1})}{&&v}}{\D v}
      && = \Diff[B]{\App{f_2}{\Update*[A]{v}{\D v}}&&}{\App{f_1}{v}}\qedAligned
  \end{alignat*}
\end{definition}

@@

  Let $a_1 \in A$ be an arbitrary value with a corresponding
  change $\D a \in \Change[A]{a}$, and let $a_2$ be
  $\Apply{\D a}{a_1}$, then
  \begin{align*}
  & \Apply[B]
      {\App{\App{\Diff*[A \to B]{f_2}{f_1}}{a_1}}{\D a}}
      {\App{f_1}{a_1}}\\
  & \quad = \Apply[B]
               {\Diff*[B]
                 {\App{f_2}{a_2}}
                 {\App{f_1}{a_1}}}
               {\App{f_1}{a_1}}\\
  & \quad = \App{f_2}{a_2}\\
  & \quad = \Apply[B]
              {\Diff*[B]
                {\App{f_2}{a_2}}
                {\App{f_1}{a_2}}}
              {\App{f_1}{a_2}}\\
  & \quad = \Apply[B]
              {\Diff*[B]
                {\App{f_2}{\Apply*{\NilC[B]{a_2}}{a_2}}}
                {\App{f_1}{a_2}}}
              {\App{f_1}{a_2}}\\
  & \quad = \Apply[B]
              {\App{\App{\Diff*[A \to B]{f_2}{f_1}}{a_2}}{\NilC{a_2}}}
              {\App{f_1}{a_2}}
  \end{align*}
  by
  \cref{def:function-changes:diff,def:update-diff,thm:update-nil}.
\end{optionalproof}

@@

  Let $f_1, f_2 \in A \to B$ be arbitrary functions. We show that
  $\Apply[A \to B]{\Diff*[A \to B]{f_2}{f_1}}{f_1}$ is
  extensionally equal to $f_2$. Let $a \in A$ be an arbitrary
  value, then
  \begin{align*}
    & \App{\Apply*[A \to B]{\Diff*[A \to B]{f_2}{f_1}}{f_1}}{a}\\
    & \quad = \Apply[B]
                {\App{\App{\Diff*[A \to B]{f_2}{f_1}}{a}}{\NilC[A]{a}}}
                {\App{f_1}{a}}\\
    & \quad = \Apply[B]
                {\Diff*[B]{\App{f_2}{\Apply*[A]{a}{\NilC[A]{a}}}}{\App{f_1}{a}}}
                {\App{f_1}{a}}\\
    & \quad = \Apply[B]
                {\Diff*[B]{\App{f_2}{a}}{\App{f_1}{a}}}
                {\App{f_1}{a}}\\
    & \quad = \App{f_2}{a}
  \end{align*}
  by the definitions of $\APPLY[A \to B]$ and $\DIFF[A \to B]$,
  \cref{thm:update-nil} for the change structure $\ChangeStruct{A}$ and
  \cref{def:update-diff} for the change structure $\ChangeStruct{B}$.
\end{optionalproof}

@@

  Let $f$, $a$, $\D f$ and $\D a$ be arbitrary, as in the statement. Then
  \begin{align*}
    & \App{\Apply*[A \to B]{\D f}{f}}{\Apply*[A]{\D a}{a}}\\
    & \quad = \Apply[B]{\App{\App{\D f}{\TheNewValue}}{\NilC{\TheNewValue}}}{\App{f}{\TheNewValue}}\\
    & \quad = \Apply[B]{\App{\App{\D f}{a}}{\D a}}{\App{f}{a}}
  \end{align*}
  by
  \cref{def:function-changes:update,def:function-changes:validity}
  as required.
\end{optionalproof}

@@

One can also compute a pointwise change from a function change:
%In particular, a pointwise change can be obtained from a function
%change by substituting to $da$ a nil change $\NilC{a}$. The result of
%$\App{\App{f'}{\Old{a}}}{\NilC{a}}$ is also a nil change (by
%\cref{thm:deriv-nil}), and $\New{a} = \Old{a}$, so we obtain:
\[
  \Update{\App {f} {a}}{\App{\App{\D f}{a}}{\NilC{a}}}
= \App{f}{a} \UPDATE \App{\nabla f}{a}
\]

%If we substitute $\App{\nabla f}{\New{a}}$ away in the equation before, we obtain the equality:
%\begin{align*}
%  & \Update{\App {\Old{f}} {\Old{a}}}{\App{\App{\D f}{\Old{a}}}{\D a}} \\
%= & \App{\Old{f}}{\New{a}} \UPDATE \App{\nabla f}{\New{a}} \\
%= & \App{\Old{f}}{\New{a}} \UPDATE \App{\App{\D f}{\New{a}}}{\NilC{\New{a}}}
%\end{align*}
%
%The above discussion was informal. To formalize it, we must
%proceed in the opposite way: we incorporate this equality in the
%definition of function changes, define $\UPDATE$ and $\DIFF$ for
%changes, and only then we can finally state and prove the
%incrementalization theorem, since the formal statement depends on
%the definition of change structures.

@@

\begin{theorem}[Nil changes are derivatives]
  \label{thm:nil-is-derivative}
  Given change structures $\ChangeStruct{A}$ and $\ChangeStruct{B}$ and a function $f \in A \to B$,
  the change $\NilC[A \to B]{f}$ is the derivative $f'$ of $f$.
\end{theorem}

\begin{optionalproof}
  Let $a \in A$ be an arbitrary value with a corresponding change
  $\D a \in \Change[A]{a}$. Then
  \begin{align*}
    & \App{f}{\Apply*[A]{\D a}{a}}\\
    & \quad = \App{\Apply*[A \to B]{\NilC[A \to B]{f}}{f}}{\Apply*[A]{\D a}{a}}\\
    & \quad = \Apply[B]{\App{\App{\NilC[A]{f}}{a}}{\D a}}{\App{f}{a}}
  \end{align*}
  holds by \cref{thm:update-nil,thm:incrementalization}, as
  required for derivatives by \cref{def:derivatives}.
\end{optionalproof}

@@

\input{pldi14/fig-function-change}
\end{oldSec}
